Rule Synthesis for Rule-based Intrusion Detectors

### 2020-05-18 paper/paper.tex
pass

\begin{abstract}
Network Intrusion Detection Systems (\nids{}) are popular tools to
defend against network attacks. These systems monitor the network
traffic and flag suspicious network behavior. Rule-based \nids\ do
that by checking the network traffic against a set of rules, which
become obsolete as attackers learn new strategies to circumvent
existing defenses. This paper proposes \tname{}, a novel approach to
synthesize rules for \textit{rule-based} \nids. \tname\ leverages positive and
negative traffic 
\rui{positive and negative traffic are first introduced here, withuot being explaiend.} 
and examples of trusted rules to create
rules not previously captured. \tname{} is organized as a pipeline of three components that
1)~create an over-specified seed rule, 2)~derive plausible rules from
the seed, and 3)~rank plausible rules. Experimental results provide
evidence that \tname\ is promising. \rui{the example is interesting, but this gives the 
feeling that the work is too preliminary} For example, it was
able to synthesize the correct rule in all cases we analyzed and
ranked the correct rule among the first five rules of the reported
ranking in \percTopFiveRanking\ of the cases.
\end{abstract}

@@

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions}. A
\emph{rule-based intrusion detector}\MyComment{\footnote{\aka\ signature-based
  intrusion detector.}} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. A rule
describes a pattern in the traffic. The observation of that pattern in
the traffic indicates an attempt of a malicious individual
to attack a local network. The example rule prescribes a method to
capture an Active Reconnaissance attack for discovering accessible
hosts in a network (see Section~\ref{sec:active-recon}).

Open-source \nids{}, like Snort~\cite{snort} and
Suricata~\cite{suricata}, provide rulesets for many existing
attacks. Unfortunately, attackers actively look for new mechanisms to
\begin{wrapfigure}[8]{r}{0.21\textwidth}
    \vspace{-2ex}  
    \centering
    \scalebox{0.9}{  
      \begin{tikzpicture}
        \begin{axis}[
            width = 0.3\textwidth,
            height = 0.2\textwidth,            
            bar width=4pt,
            scaled ticks=false,
            ymajorticks=false,
            ybar stacked,
            enlargelimits=0.09,
            clip=true,
            legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
            xtick=data,
            xticklabel style={rotate=45, font=\footnotesize},
            tick pos=left,
            nodes near coords,
            every node near coord/.append style={font=\footnotesize},
            nodes near coords align={vertical},
            symbolic x coords={Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec},]
            \addplot coordinates {(Jan, 114) (Feb, 155) (Mar, 117) (Apr, 159) (May, 110) (Jun, 231) (Jul, 122) (Aug, 167) (Sep, 107) (Oct, 225) (Nov, 148) (Dec, 127)};
        \end{axis}
      \end{tikzpicture}
    }
    \vspace{-5ex}
    \caption{\label{fig:distribution-rules-per-month}2019 new rules.}
\end{wrapfigure}
exploit systems through the
network. Figure~\ref{fig:distribution-rules-per-month} shows the
number of rules added per month in the year of 2019 to the ``Emerging
Threats Open Ruleset''~\cite{emerging-threats-open}, a popular public
data set of \suri\ rules. A closer observation on the
changelogs of this ruleset~\cite{emerging-threats-changelogs} shows that new rules are
often created.  Rulesets need to be continuously revised because the
community finds and advertises vulnerabilities, which eventually can be exploited by attackers. 
Unfortunately, creating these rules is
time-consuming and error-prone \rui{is there a citation we could add here?}. The language for describing rules is
non-trivial (see Section~\ref{sec:example-suricata-rules}) and testing
correctness of a rule requires data and careful inspection.

@@

This paper proposes \tname{} (for \textbf{Sy}nthesis of
Su\textbf{ri}cata R\textbf{u}le\textbf{s}), a technique to
automatically synthesize rules for rule-based \nids. Our goal is to
facilitate the creation of rules and, consequently, help system
administrators to harden the protection against network attacks.
\tname{} leverages different sources of information to synthesize
rules for a given attack. It uses 1) positive examples, \ie{},
scenarios that reproduce the attack, 2) negative examples in the form
of a data set of benign messages, and 3) a data set of human-written
trusted rules. The positive and negative examples constrain the search
space of rules that can be used to capture the attack. Intuitively,
positive examples indicate what should be captured whereas negative
examples indicate what should not be captured. It is worth noting that
vast amounts of negative data
exist~\cite{tcpreplay,stratosphere-normal}, but positive data is
scarce as it is not common to publish attacks openly on the
web. Classical machine learning techniques \emph{are not directly
  applicable in this context} for two main reasons: 1)~they require a
large amount of training data to build models (recall that positive
data is scarce) and 2)~they typically have limited support to string
and structured data, which are prevalent in this domain.

@@

We evaluated \tname\ against 14 attacks of various kinds, such as

\rui{add few examples}. The observed results are encouraging. For 
example, \tname\ ranked the correct rule among the top five rules in 
\percTopFiveRanking\ of the cases and ranked the correct rule among 
the top ten rules in 11 of the 14 cases. \rui{to discuss, I find 
this a bit weak.}

>>>>

\begin{abstract}
Network Intrusion Detection Systems (\nids{}) are popular tools to
defend against network attacks. These systems monitor the network
traffic and flag suspicious network behavior. Rule-based \nids\ do
that by checking the network traffic against a set of rules, which
become obsolete as attackers learn new strategies to circumvent
existing defenses. This paper proposes \tname{}, a novel approach to
synthesize rules for \textit{rule-based} \nids. \tname\ leverages
malicious (positive) and benign (negative) traffic 
%\rui{positive and negative traffic are first introduced here, withuot being explaiend.} 
and examples of trusted rules to create
rules for new attacks. \tname{} is organized as a pipeline of three components that
1)~create an over-specified seed rule, 2)~derive plausible rules from
the seed, and 3)~rank plausible rules. Experimental results provide
evidence that \tname\ is promising.
 \rui{the example is interesting, but this gives the 
feeling that the work is too preliminary} 
For example, it was
able to synthesize the correct rule in all cases we analyzed and
ranked the correct rule among the first five rules of the reported
ranking in \percTopFiveRanking\ of the cases.
\end{abstract}

@@

This paper focuses on rule-based \nids{}, which are very popular in
industry~\cite{proofpoint-etpro,snort-rule-subscriptions}. A
\emph{rule-based intrusion detector}\MyComment{\footnote{\aka\ signature-based
  intrusion detector.}} checks if the network traffic matches a fixed
set of rules. Figure~\ref{fig:pingscan-example} shows an example rule
of Suricata~\cite{suricata}, a popular open-source \nids{}. A rule
describes a pattern in the traffic. The observation of that pattern in
the traffic indicates an attempt of a malicious individual
to attack a local network. This particular rule prescribes a method to
capture an Active Reconnaissance attack for discovering accessible
hosts in a network (see Section~\ref{sec:active-recon}).

Open-source \nids{}, like Snort~\cite{snort} and
Suricata~\cite{suricata}, provide rulesets for many existing
attacks. Unfortunately, attackers actively look for new mechanisms to
\begin{wrapfigure}[8]{r}{0.21\textwidth}
    \vspace{-2ex}  
    \centering
    \scalebox{0.9}{  
      \begin{tikzpicture}
        \begin{axis}[
            width = 0.3\textwidth,
            height = 0.2\textwidth,            
            bar width=4pt,
            scaled ticks=false,
            ymajorticks=false,
            ybar stacked,
            enlargelimits=0.09,
            clip=true,
            legend style={at={(0.5,-0.15)},anchor=north,legend columns=-1},          
            xtick=data,
            xticklabel style={rotate=45, font=\footnotesize},
            tick pos=left,
            nodes near coords,
            every node near coord/.append style={font=\footnotesize},
            nodes near coords align={vertical},
            symbolic x coords={Jan, Feb, Mar, Apr, May, Jun, Jul, Aug, Sep, Oct, Nov, Dec},]
            \addplot coordinates {(Jan, 114) (Feb, 155) (Mar, 117) (Apr, 159) (May, 110) (Jun, 231) (Jul, 122) (Aug, 167) (Sep, 107) (Oct, 225) (Nov, 148) (Dec, 127)};
        \end{axis}
      \end{tikzpicture}
    }
    \vspace{-5ex}
    \caption{\label{fig:distribution-rules-per-month}2019 new rules.}
\end{wrapfigure}
exploit systems through the
network. Figure~\ref{fig:distribution-rules-per-month} shows the
number of rules added per month in the year of 2019 to the ``Emerging
Threats Open Ruleset''~\cite{emerging-threats-open}, a popular public
data set of \suri\ rules. A closer observation on the
changelogs of this ruleset~\cite{emerging-threats-changelogs} shows that new rules are
often created.  Rulesets need to be continuously revised as the
community finds and advertises vulnerabilities, which eventually can be exploited by hackers. 
Unfortunately, creating these rules is
time-consuming and error-prone \rui{is there a citation we could add here?}. The language for describing rules is
non-trivial (see Section~\ref{sec:example-suricata-rules}) and testing
correctness of a rule requires data and careful inspection.

@@

\sloppy This paper proposes \tname{} (for \textbf{Sy}nthesis of
Su\textbf{ri}cata R\textbf{u}le\textbf{s}), a technique to
automatically synthesize rules for rule-based \nids. Our goal is to
facilitate the creation of rules and, consequently, help system
administrators to harden the protection against network attacks.
\tname{} leverages different sources of information to synthesize
rules for a given attack. It uses 1) positive examples, \ie{},
scenarios that reproduce the attack, 2) negative examples in the form
of a data set of benign messages, and 3) a data set of human-written
trusted rules. The positive and negative examples constrain the search
space of rules that can be used to capture the attack. Intuitively,
positive examples indicate what should be captured whereas negative
examples indicate what should not be captured. It is worth noting that
vast amounts of negative data
exist~\cite{tcpreplay,stratosphere-normal}, but positive data is
scarce as it is not common (and considered unethical~\Mar{look for
  citation for ethics code on this}) to publish attacks on the
web\Mar{is there a term for non-deep web?}. Classical machine learning techniques \emph{are not directly
  applicable in this context} for two main reasons: 1)~they require a
large amount of training data to build models (recall that positive
data is scarce) and 2)~they typically have limited support to string
and structured data, which are prevalent in this domain.

@@

We evaluated \tname\ against \Mar{@Lucas, please look for constant
  numbers as this *in text* and use macros instead. We do not want to
  leave inconsistent data in the paper!}14 attacks of various kinds, such as \Fix{...}\rui{add few examples}. The observed results are encouraging. For 
example, \tname\ ranked the correct rule among the top five rules in 
\percTopFiveRanking\ of the cases and ranked the correct rule among 
the top ten rules in 11 of the 14 cases. \rui{to discuss, I find 
this a bit weak.}
